Variation in gaze understanding across the life span: A process-level perspective

### 2023-10-04 paper/manuscript.Rmd
ready for first round of co-author feedback

abstract: |
  Observing eye gaze is fundamental for many social-cognitive abilities, for example, when judging what another agent can or cannot know. While the emergence of gaze following - on a group level - has been thoroughly studied, we know little about (a) the developmental trajectory beyond infancy and (b) the sources of individual differences. In Study 1, we examine gaze understanding across the lifespan (*N* = 471 3- to 80-years-olds from a mid-sized German city). We found a steep learning curve during preschool years, in which children become more precise in locating the attentional focus of an agent. Precision levels then stay comparably stable throughout adulthood with a minor decline toward old age. In Study 2, we formalize the process of gaze understanding in a computational cognitive model that allows us to conceptualize individual differences in a psychologically meaningful way (*N* = 60 3- to 5-year-olds, 50 adults). According to our model, participants estimate pupil angles with varying levels of precision based on observing the pupil location within the agent's eyes. In Study 3, we test two implications of our gaze model, namely that gaze understanding should be related to perspective-taking abilities and vector estimation in non-social settings (*N* = 102 4- to 5-year-olds). We found that gaze understanding is associated with both of these abilities but not with other Theory of Mind tasks. This work illustrates how the combination of reliable measurement instruments and formal theoretical models allows us to explore the in(ter)dependence of core social-cognitive processes in greater detail.
keywords          : "social-cognitive development, theory of mind, gaze cues, individual differences, cognitive modeling, lifespan"
wordcount         : "xxx max 8000"
floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes 
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_word
header-includes   :
  - \usepackage{setspace}
  - \captionsetup[figure]{font={stretch=1}}
editor_options    : 
  markdown: 
    wrap: sentence
bibliography      : "../../references.bib"
---

@@

At first glance, the debate about children’s gaze following abilities seems to be settled: at 12 months of age, children are already competent gaze followers. On further view, however, some methodological choices in the experimental paradigms as well as the scope of infants’ abilities seem to offer room for further exploration. We see the following challenges: (1) Studies focus on infancy, leaving aside further development in childhood, adolescence and adulthood; (2) Gaze cues are often accompanied by head movements in the same direction [see @delbianco2019developmental], which makes it hard to isolate the referential information extracted from the eyes; (3) Infants’ looking times are usually used as the dependent variable and it is unclear if and how this would translate into active behavior; (4) Presenting two objects drastically limits the complexity of the visual input, especially when compared to our daily, visually cluttered environment; (5) Analyses focus on group-level averages, and we overlook variation in the sample because we cannot measure it with dichotomous outcome variables. In the following paragraphs, we will turn to each of these arguments and point out gaps in our understanding of the development of gaze following.

First, while the emergence of gaze following has been well established, less is known about the developmental trajectory beyond infancy. One possibility is that the ability to follow gaze does not improve further once it emerges. Yet, most - if not all - cognitive abilities develop throughout childhood [e.g., working memory, @gathercole2004structure]. Similarly, visual processing appears to improve with age. First longitudinal evidence points to a development in the infant's individual joint attention accuracy and latency [@gredeback2010development]. It seems likely that children fine-tune the performance of the already existing skills - presumably while using them in social interactions. To capture further development in gaze understanding, we therefore conduct our study across the lifespan.

Second, we see in traditional gaze following studies that the experimenter often shifts their eyes and head in synchrony [e.g., @behne2005oneyearolds; @okumura2017individual] so that children's gaze following could be interpreted as following face direction alone. A handful of studies approached this potential confound by separately manipulating head and eye movement. @michel2021effects found that gaze following in four-month-olds was likely driven by another's head instead of their eye movement. @brooks2002importance implemented a comparison between eye and head orientation and found that 14-month-olds were sensitive to open versus closed eyes [see also @moore1997role]. Results of @corkum1995development, @lempers1979young, and @lempers1977development suggest that infants, at least until 19 months, struggle when eye and head direction diverge. From farther distance, body or face orientation can act as more salient cues to determine another's area of attention. However, eye direction can indicate a more precise location of focus [@emery2000eyes], and allows us to anticipate likely future actions [@zohary2022gaze; @friesen2011gaze]. Clearly, it is important to differentiate between head and gaze following. In this study, we focus on subtle gaze cues and isolate eye movement alone.

Third, let us consider the classical response measure of collecting looking times. Though previous studies suggest that young infants can align their visual attention to another’s line of sight, it does not necessarily include understanding the intentions or the focus of the agent. Infants could simply align their orientation without processing what exactly the other is seeing [cf. Butterworth & Jarrett's ecological and geometric mechanism, @butterworth1991minds]. In fact, one might ask if such an alignment reflects understanding of visual perspectives at all, because the “target” or “object of representation” is not necessarily specified [@perner2003perspective, p. 358]. To fulfill this criterion, @moll2006level used an active behavioral choice as a response measure. Participants had to search for the target object, which was either visually accessible or behind a barrier. 24-month-olds succeeded in the task, while 18-month-olds did not. In an object choice study with two containers, @behne2005oneyearolds found that already 14-month-olds used an experimenter's gaze; however, this was again confounded with a congruent face movement [cf. @povinelli1997exploitation]. In this paper, we assess how subtle gaze cues guide participants' behavior in actively locating a target.

Finally, presenting two objects and focusing on group-level analyses masks individual differences between children in their gaze following abilities. Measuring individual differences in basic aspects of social cognition is especially important to understand the underlying processes and to quantify the impact of environmental influences and other cognitive abilities [@birch2017perspectives; @delbianco2019developmental]. In this study, we measure participants' gaze understanding continuously by using a task which is designed to capture individual-level variation [@prein2023tango].

@@

The present study has two main goals: first, we study the development of gaze understanding beyond infancy on a group level (Study 1). Our goal was not to establish the youngest age at which children understand gaze cues. Rather, we wanted to examine how this ability changes with age. Even though we find expected developmental patterns - rapid improvement in early childhood; some decline toward old age - we observed substantial variation between individuals at all ages. Our second goal is to provide a substantive process-level theory of gaze understanding (and individual differences therein) in the form of a computational model. We formalize gaze understanding as a form of vector estimation and show that our model explains empirical data (Study 2). Furthermore, we test two empirical predictions that follow from the model, namely that gaze understanding is related to non-social vector estimation and visual perspective-taking (Study 3).

@@

We aimed to assess the developmental trajectory of gaze understanding across the lifespan. The study was pre-registered at: <https://osf.io/snju6> (child sample) and <https://osf.io/6yjz3> (adult sample). The study obtained ethical clearance by the MPG Ethics commission Munich, Germany, falling under an umbrella ethics application (Appl. No. 2021_45), and was approved by an internal ethics committee at the Max Planck Institute for Evolutionary Anthropology. The research adheres to the legal requirements of psychological research with children in Germany. Data was collected between May 2021 and April 2023.

@@

To estimate the effect of age on gaze understanding, we fit GLMMs that make different assumptions about the developmental trajectory, modeling the relationship as linear, quadratic, or cubic. In addition, we applied a Gaussian Process model [@burkner2017brms] and examined nonlinearity. The models predicted imprecision by age (continuous), aggregated across trials and modeled as a `lognormal` distribution.[^1] The unit of imprecision was counted in widths, i.e., a participant with an imprecision of 1 clicked on average one balloon width to the left or right of the true target center. We inspected the posterior distributions (mean and 95% Credible Interval (CrI)) for the age estimates and compared models using leave-one-out cross-validation (LOO), the widely applicable information criterion (WAIC), and model weights [@vehtari2017practicala].

@@

In our model comparison, we found clear evidence for a non-linear development in gaze understanding across the lifespan. Compared against linear, quadratic, and cubic models, the Gaussian Process model showed the best model fit, demonstrated by the greatest model weight and smallest WAIC values. For the imprecision in gaze understanding, the standard deviation (*b* = `r lifespan_gp$estimate[lifespan_gp$param == "sdgp(gpage_centered)"]`, 95% CrI [`r lifespan_gp$l_95_percent_ci[lifespan_gp$param == "sdgp(gpage_centered)"]`; `r lifespan_gp$u_95_percent_ci[lifespan_gp$param == "sdgp(gpage_centered)"]`]) and visual inspection indicated nonlinearity (see Figure \@ref(fig:fig1)).

@@

Our lifespan study showed that gaze understanding develops throughout childhood, and variation between individuals appears in all age groups. The TANGO has previously been shown to reliably capture inter-individual differences in gaze understanding [@prein2023tango]. The variation between participants is thus likely genuine and not due to random noise. Now, we aim to understand developmental change and variation across participants on a process level. We present a theory of gaze understanding that explains how children process the available information (i.e., the agent's eyes) to make inferences about the agent's line of sight and its attentional focus, which leads them to identify the target position. We formalize this inference process in a computational cognitive model.

Computational modeling frameworks allow researchers to establish mechanistic explanations of psychological phenomena [@grahek2021anatomy]. As formal, mathematical accounts of the psychological process in question, they force researchers to accurately and comprehensively state all their underlying assumptions [@simmering2010dialogue]. In addition, models can be used to simulate behavior and form testable predictions. The expected patterns can then, in turn, be compared to the empirically observed behavior. @moore2008development have already argued for systematic, empirical approaches in combination with formal modeling to arrive at a deeper understanding of gaze following.

@@

Our model quantifies a participant's cognitive ability to follow gaze by inverting a probabilistic process that generates the participant's clicks from observing the eyes of the agent. It is formally defined as:

@@

[^3]: This model mirrors the logic of the TANGO programming code. In the online experiment, we read out the center point coordinates of the target and the agent's eyeball (i.e., the SVG coordinates). We then calculate a line between these two points: this is our gaze vector (acting in the functionally same way as a pupil angle). Now, knowing the eyeball radius, we calculate the point of intersection at which the gaze vector meets the eyeball boundary. Finally, the agent's pupil moves from the center of the eyeball along the gaze vector to the intersection point. This way, the agent is animated to "look at" the target. In the gaze model, we assume participants go through these steps in reverse order.

Based on our verbal task instructions, we assume that participants (1) expect the agent's looks to be directed at the target, and (2) to click on the coordinate they estimate the agent to look at. Consequently, we do not assume that participants' clicks are noisy in any way but that they click on the screen location where they genuinely think the target is (and that the agent is looking at).

The true eye angles ($\alpha_l$ and $\alpha_r$) cannot be directly observed and have to be estimated based on the position of the pupils within the eyes, resulting in approximate values ($\hat{\alpha_l}$ and $\hat{\alpha_r}$). We presume this estimation to be a noisy process. Thus, we conceptualize the development of the cognitive ability to follow gaze as a reduction of noise in the estimates (i.e., an increased certainty about the pupil angles).

Any clicked value of $x_c$ implies a "matched pair" of the estimated pupil angles $\hat{\alpha_l}$ and $\hat{\alpha_r}$, with the property that lines extended along those two angles meet at the precise location of the target. As a consequence, we can rewrite the likelihood function of the model above:

@@

The second term of the right-hand side equation above, $P(x_c)$, is a prior over potential target locations, which we assume to be skewed towards the screen center: We anticipate that participants have an a priori expectation that the target will land close to the middle, partly because the target was last visible in the screen center before disappearing behind the hedge, and because the agent is located centrally on the screen. We estimate the strength of this center bias (i.e., the standard deviation of a Normal distribution around the center of the screen) based on the data: $P(x_c) \sim \mathcal{N}(960, \sigma^p)$.

The width of this distribution is defined by $\sigma^p$. We assumed that the center bias changes with age and we therefore estimated $\sigma^p$ via a linear regression as a function of the participant's age ($age_i$): $\sigma^p = \beta_0^{\sigma^p} + age_i \cdot \beta_1^{\sigma^p}$. Therefore, the participant-specific value for $P(x_c)$ was constrained by the performance in the TANGO and the participant's age.

The main inferential task for the participant lies in estimating the pupil angles, i.e., sampling from the first term of the right-hand side equation above, $P(\hat{\alpha_l}, \hat{\alpha_r} | \alpha_l, \alpha_r, \theta)$. For this, we assume that the pair of estimated pupil angles are sampled from a probability distribution which is the product of two Normal distributions of equal variance, $\sigma_v$, centered on the true pupil angles:

@@

Broadly summarizing: Participants are assumed to observe the pupil locations and the center of the agent's eyes. Connecting these two point estimates as a line yields the unique vector that extends from the center of the agent's eyeball through the center of the pupil to the attentional focal point. Taking the angle between this vector and a line pointing vertically to the ground yields the pupil angle. Participants are assumed to sample from Normal distributions centered around the true pupil angle. They are assumed to do this independently for the left and right eye and then integrate the information to estimate the target's location.

As $\sigma_v$ determines the level of accuracy with which the participant estimates the pupil angles, it is the component that defines $\theta$. When $\sigma_v$ is very small (i.e., the distribution around the pupil angle is narrow), clicks far away from the target are unlikely, as these would require estimated pupil angles very different from the true pupil angles.\ When $\sigma_v$ is very large (i.e., the distribution around the pupil angle is wide), almost any pupil angles may be sampled, corresponding to a roughly uniform distribution over click coordinates. We expect $\sigma_v$ to vary between individuals. Consequently, individuals differ in the level of precision with which they can locate the target based on observing the agent's eyes.

The shape of the $P(x_c | \alpha_l, \alpha_r, \theta)$ distribution leads to a testable group-level prediction. As the pupil location varies, a fixed amount of uncertainty around the pupil angle corresponds to a varying degree of uncertainty in the estimated target location (see Figure \@ref(fig:fig2)B & C). When the agent directs their gaze toward the very left or right side, the distribution around the pupil angle from which participants sample is comparatively wider than when the agent gazes centrally to the ground in front of them. For illustrative purposes, imagine a similar phenomenon: pointing a torch light to a flat surface on the ground. When one points the light cone directly at the surface, the light beam is concentrated in a clearly defined, small, symmetric area. When one points the light cone further away from oneself (shining at an angle), the light from one half of the cone must travel further to reach the surface than the light from the other half, resulting in an asymmetric light pattern. As the angle increases, the light is spread over a wider area, and the surface is illuminated less evenly. Consequently, for the same $\sigma_v$, the further out a target coordinate lies, the wider and less symmetric the distribution. This increases both the variance and the bias in a participant's estimate of the agent's attentional focus, resulting in a decreased performance in the task. As $\sigma_v$ decreases and the cone narrows, the extent to which performance varies at different angles decreases. Therefore, our gaze model predicts that our trials vary in difficulty: participants should be more imprecise in locating the target the further out it lands, resulting in a U-shaped pattern.[^4] If our data matches the pattern of this model prediction, this can act as evidence for the gaze model. Therefore, our gaze model provides a quantitative theory of gaze understanding with testable predictions. In the following, we test these predictions in children and adults.

@@

Our goal was to describe the inferential process of gaze understanding. We quantified how well our gaze model explained the gaze understanding process in two ways. First, we used correlations to quantify how well the gaze model is able to recover the data. For this, we aggregated the model predictions and the data for each target bin and age group (3-, 4-, 5-years-olds), and computed the coorrelation between predictions and the data. Second, we compared the predictions of our gaze model to two - very simple - alternative models that assume that participants do not rely on the agent's gaze at all: a random guessing model and a center bias model. The random guessing model assumed that participants randomly clicked on the screen and was implemented as sampling from a uniform distribution over all possible coordinates, $\mathcal{U}(0, 1920)$. The center bias model assumed that participants always click at the screen center and was implemented as sampling from a Normal distribution with the the center of the screen as the mean, and one balloon width as the standard deviation, $\mathcal{N}(960, 160^2)$. Note that the center bias model also predicts that imprecision should be higher for targets further out on the screen, resulting also in a U-shaped pattern. However, the shape of this pattern is different from that of the gaze model [TODO: in what way?!]. All cognitive models were implemented in WebPPL [@goodman2014design].

@@

```{r gazemodel_results}
gazemodel_comparison_bf <- readRDS("../data/gazemodel-kids-comparison-bf.rds")
gazemodel_corrs <- readRDS("../saves/gazemodel_correlations.rds")
```

>>>>

abstract: |
  Observing eye gaze is fundamental for many social-cognitive abilities, for example, when judging what another agent can or cannot know. While the emergence of gaze following - on a group level - has been thoroughly studied, we know little about (a) the developmental trajectory beyond infancy and (b) the sources of individual differences. In Study 1, we examined gaze understanding across the lifespan (*N* = 471 3- to 80-years-olds; children from a mid-sized German city; international remotely tested adults). We found a steep learning curve during preschool years, in which children become more precise in locating the attentional focus of an agent. Precision levels then stayed comparably stable throughout adulthood with a minor decline toward old age. In Study 2, we formalized the process of gaze understanding in a computational cognitive model that allowed us to conceptualize individual differences in a psychologically meaningful way (*N* = 60 3- to 5-year-olds, 50 adults). According to our model, participants estimate pupil angles with varying levels of precision based on observing the pupil location within the agent's eyes. In Study 3, we tested two implications of our gaze model, namely that gaze understanding should be related to perspective-taking abilities and vector estimation in non-social settings (*N* = 102 4- to 5-year-olds). We found that gaze understanding is associated with both of these abilities but less so with other Theory of Mind tasks. This work illustrates how the combination of reliable measurement instruments and formal theoretical models allows us to explore the in(ter)dependence of core social-cognitive processes in greater detail.
keywords          : "social-cognitive development, theory of mind, gaze cues, individual differences, cognitive modeling, lifespan"
wordcount         : "xxx max 8000"
floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes 
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_word
header-includes   :
  - \usepackage{setspace}
  - \captionsetup[figure]{font={stretch=1}}
editor_options    : 
  markdown: 
    wrap: sentence
bibliography      : "../../references.bib"
---

@@

At first glance, the debate about children’s gaze following abilities seems to be settled: at 12 months of age, children are already competent gaze followers. On further view, however, some methodological choices in the experimental paradigms as well as the scope of infants’ abilities seem to offer room for further exploration. We see the following challenges: (1) Studies focus on infancy, leaving aside further development in childhood, adolescence and adulthood; (2) Gaze cues are often accompanied by head movements in the same direction [see @delbianco2019developmental], which makes it hard to isolate the referential information extracted from the eyes; (3) Infants’ looking times are usually used as the dependent variable and it is unclear if and how this would translate into active behavior; (4) Presenting two objects drastically limits the complexity of the visual input, especially when compared to our daily, visually cluttered environment; (5) Analyses focus on group-level averages, and we overlook variation in the sample because we cannot measure it with dichotomous outcome variables. In the following paragraphs, we will turn to each of these arguments and note how we addressed them consequently in our study.

First, while the emergence of gaze following has been well established, less is known about the developmental trajectory beyond infancy. One possibility is that the ability to follow gaze does not improve further once it emerges. Yet, most - if not all - cognitive abilities develop throughout childhood [e.g., working memory, @gathercole2004structure]. Similarly, visual processing appears to improve with age. First longitudinal evidence points to a development in the infant's individual joint attention accuracy and latency [@gredeback2010development]. It seems likely that children fine-tune the performance of the already existing skills - presumably while using them in social interactions. To capture further development in gaze understanding, we therefore conducted our Study 1 across the lifespan.

Second, we see in traditional gaze following studies that the experimenter often shifts their eyes and head in synchrony [e.g., @behne2005oneyearolds; @okumura2017individual] so that children's gaze following could be interpreted as following face direction alone. A handful of studies approached this potential confound by separately manipulating head and eye movement. @michel2021effects found that gaze following in four-month-olds was likely driven by another's head instead of their eye movement. @brooks2002importance implemented a comparison between eye and head orientation and found that 14-month-olds were sensitive to open versus closed eyes [see also @moore1997role]. Results of @corkum1995development, @lempers1979young, and @lempers1977development suggest that infants, at least until 19 months, struggle when eye and head direction diverge. From farther distance, body or face orientation can act as more salient cues to determine another's area of attention. However, eye direction can indicate a more precise location of focus [@emery2000eyes], and allows us to anticipate likely future actions [@zohary2022gaze; @friesen2011gaze]. Clearly, it is important to differentiate between head and gaze following. In our three studies, we focused on subtle gaze cues and isolated eye movement alone.

Third, let us consider the classical response measure of collecting looking times. Though previous studies suggest that young infants can align their visual attention to another’s line of sight, it does not necessarily include understanding the intentions or the focus of the agent [cf. @aslin2007look]. Infants could simply align their orientation without processing what exactly the other is seeing [cf. Butterworth & Jarrett's ecological and geometric mechanism, @butterworth1991minds]. In fact, one might ask if such an alignment reflects understanding of visual perspectives at all, because the “target” or “object of representation” is not necessarily specified [@perner2003perspective, p. 358]. To fulfill this criterion, @moll2006level used an active behavioral choice as a response measure. Participants had to search for the target object, which was either visually accessible or behind a barrier. 24-month-olds succeeded in the task, while 18-month-olds did not. In an object choice study with two containers, @behne2005oneyearolds found that already 14-month-olds used an experimenter's gaze; however, this was again confounded with a congruent face movement [cf. @povinelli1997exploitation]. In this paper, we assessed how subtle gaze cues guided participants' behavior in actively locating a target.

Finally, presenting two objects and focusing on group-level analyses masks individual differences between children in their gaze following abilities. Measuring individual differences in basic aspects of social cognition is especially important to understand the underlying processes and to quantify the impact of environmental influences and other cognitive abilities [@birch2017perspectives; @delbianco2019developmental]. In our studies, we measured participants' gaze understanding continuously by using a task which has been designed to capture individual-level variation [@prein2023tango].

@@

The present study had two main goals: first, we studied the development of gaze understanding beyond infancy on a group level (Study 1). Instead of capturing the youngest age at which children can follow gaze, we wanted to examine how this ability changes with age. Even though we found expected developmental patterns - rapid improvement in early childhood; some decline toward old age - we observed substantial variation between individuals at all ages. Our second goal was to provide a substantive process-level theory of gaze understanding (and individual differences therein) in the form of a computational model. We formalized gaze understanding as a form of vector estimation and showed that our model explained empirical data (Study 2). Furthermore, we tested two empirical predictions that followed from the model, namely that gaze understanding is related to non-social vector estimation and visual perspective-taking (Study 3).

@@

We aimed to assess the developmental trajectory of gaze understanding across the lifespan. The study was pre-registered at: <https://osf.io/snju6> (child sample) and <https://osf.io/6yjz3> (adult sample). The study obtained ethical clearance by the MPG Ethics commission Munich, Germany, falling under an umbrella ethics application (Appl. No. 2021_45), and was approved by an internal ethics committee at the Max Planck Institute for Evolutionary Anthropology. The research adhered to the legal requirements of psychological research with children in Germany. Data was collected between May 2021 and April 2023.

@@

To estimate the effect of age on gaze understanding, we fit GLMMs that make different assumptions about the developmental trajectory, modeling the relationship as linear, quadratic, or cubic. In addition, we applied a Gaussian Process model [@burkner2017brms] and examined nonlinearity. The models predicted imprecision by age (continuous), aggregated across trials and modeled as a `lognormal` distribution.[^1] The unit of imprecision was counted in widths, i.e., a participant with an imprecision of 1 clicked on average one balloon width to the left or right of the true target center. We inspected the posterior distributions (mean and 95% Credible Interval (CrI)) for the age estimates and compared models via model weights and the difference in expected log pointwise predictive density (ELPD) estimated using leave-one-out cross-validation (LOO) [@vehtari2017practicala].

@@

In our model comparison, we found substantial evidence for a non-linear development in gaze understanding across the lifespan. The Gaussian process model was clearly preferred over the cubic, quadratic and linear models due to the highest predictive accuracy according to the LOO ELPD estimates (with negative ELPD differences and more than twice the estimated standard error): elpd_diff between Gaussian process and cubic model = `r lifespan_mcomparison$elpd_diff[lifespan_mcomparison$model == "mcubic"]` (SE = `r lifespan_mcomparison$se_diff[lifespan_mcomparison$model == "mcubic"]`); elpd_diff between Gaussian process and quadratic model = `r lifespan_mcomparison$elpd_diff[lifespan_mcomparison$model == "mquadratic"]` (SE = `r lifespan_mcomparison$se_diff[lifespan_mcomparison$model == "mquadratic"]`); and elpd_diff between Gaussian process and linear model = `r lifespan_mcomparison$elpd_diff[lifespan_mcomparison$model == "mlinear"]` (SE = `r lifespan_mcomparison$se_diff[lifespan_mcomparison$model == "mlinear"]`); all in favor of the Gaussian process model. Morover, the Gaussian Process model showed the greatest model weight (approximating `r lifespan_mcomparison$modelweight[lifespan_mcomparison$model == "mgaussianprocess"]`). For the imprecision in gaze understanding, the standard deviation (*b* = `r lifespan_gp$estimate[lifespan_gp$param == "sdgp(gpage_centered)"]`, 95% CrI [`r lifespan_gp$l_95_percent_ci[lifespan_gp$param == "sdgp(gpage_centered)"]`; `r lifespan_gp$u_95_percent_ci[lifespan_gp$param == "sdgp(gpage_centered)"]`]) and visual inspection indicated nonlinearity (see Figure \@ref(fig:fig1)).

@@

Our lifespan study showed that gaze understanding develops throughout childhood, and variation between individuals appears in all age groups. The TANGO has previously been shown to reliably capture inter-individual differences in gaze understanding [@prein2023tango]. The variation between participants was thus likely genuine and not due to random noise. Now, we aimed to understand developmental change and variation across participants on a process level. We present a theory of gaze understanding that explains how children process the available information (i.e., the agent's eyes) to make inferences about the agent's line of sight and its attentional focus, which leads them to identify the target position. We formalized this inference process in a computational cognitive model.

Computational modeling frameworks allow researchers to establish mechanistic explanations of psychological phenomena [@grahek2021anatomy]. As formal, mathematical accounts of the psychological process in question, they force researchers to accurately and comprehensively state all their underlying assumptions [@simmering2010dialogue]. In addition, models can be used to simulate behavior and form testable predictions. The expected patterns can then, in turn, be compared to the empirically observed behavior. @moore2008development has already argued for systematic, empirical approaches in combination with formal modeling to arrive at a deeper understanding of gaze following.

@@

Our model quantified a participant's cognitive ability to follow gaze by inverting a probabilistic process that generates the participant's clicks from observing the eyes of the agent. It is formally defined as:

@@

[^3]: This model mirrors the logic of the TANGO programming code. In the online experiment, we read out the center point coordinates of the target and the agent's eyeball (i.e., the SVG coordinates). We then calculated a line between these two points: this was our gaze vector (acting in the functionally same way as a pupil angle). Now, knowing the eyeball radius, we calculated the point of intersection at which the gaze vector met the eyeball boundary. Finally, the agent's pupil moved from the center of the eyeball along the gaze vector to the intersection point. This way, the agent was animated to "look at" the target. In the gaze model, we assumed participants go through these steps in reverse order.

Based on our verbal task instructions, we assumed that participants (1) expected the agent's looks to be directed at the target, and (2) to click on the coordinate they estimated the agent to look at. Consequently, we did not assume that participants' clicks were noisy in any way but that they clicked on the screen location where they genuinely thought the target was (and that the agent was looking at).

The true eye angles ($\alpha_l$ and $\alpha_r$) cannot be directly observed and have to be estimated based on the position of the pupils within the eyes, resulting in approximate values ($\hat{\alpha_l}$ and $\hat{\alpha_r}$). We presumed this estimation to be a noisy process. Thus, we conceptualized the development of the cognitive ability to follow gaze as a reduction of noise in the estimates (i.e., an increased certainty about the pupil angles).

Any clicked value of $x_c$ implied a "matched pair" of the estimated pupil angles $\hat{\alpha_l}$ and $\hat{\alpha_r}$, with the property that lines extended along those two angles met at the precise location of the target. As a consequence, we can rewrite the likelihood function of the model above:

@@

The second term of the right-hand side equation above, $P(x_c)$, is a prior over potential target locations, which we assumed to be skewed towards the screen center: We anticipated that participants have an a priori expectation that the target will land close to the middle, partly because the target was last visible in the screen center before disappearing behind the hedge, and because the agent was located centrally on the screen. We estimated the strength of this center bias (i.e., the standard deviation of a Normal distribution around the center of the screen) based on the data: $P(x_c) \sim \mathcal{N}(960, \sigma^p)$.

The width of this distribution is defined by $\sigma^p$. For children, we assumed that the center bias changed with age and we therefore estimated $\sigma^p$ via a linear regression as a function of the child's age ($age_i$): $\sigma^p = \beta_0^{\sigma^p} + age_i \cdot \beta_1^{\sigma^p}$. Therefore, the participant-specific distribution for $P(x_c)$ was constrained by the performance in the TANGO and the child's age. For the adults, $\sigma^p$ was not age-specific.

The main inferential task for the participant lied in estimating the pupil angles, i.e., sampling from the first term of the right-hand side equation above, $P(\hat{\alpha_l}, \hat{\alpha_r} | \alpha_l, \alpha_r, \theta)$. For this, we assumed that the pair of estimated pupil angles were sampled from a probability distribution which is the product of two Normal distributions of equal variance, $\sigma_v$, centered on the true pupil angles:

@@

Broadly summarizing: Participants were assumed to observe the pupil locations and the center of the agent's eyes. Connecting these two point estimates as a line yields the unique vector that extends from the center of the agent's eyeball through the center of the pupil to the attentional focal point. Taking the angle between this vector and a line pointing vertically to the ground yields the pupil angle. Participants were assumed to sample from Normal distributions centered around the true pupil angle. They were assumed to do this independently for the left and right eye and then integrate the information to estimate the target's location.

As $\sigma_v$ determined the level of accuracy with which the participant estimated the pupil angles, it is the component that defines $\theta$. When $\sigma_v$ is very small (i.e., the distribution around the pupil angle is narrow), clicks far away from the target are unlikely, as these would require estimated pupil angles very different from the true pupil angles.\ When $\sigma_v$ is very large (i.e., the distribution around the pupil angle is wide), almost any pupil angles may be sampled, corresponding to a roughly uniform distribution over click coordinates. We expected $\sigma_v$ to vary between individuals. Consequently, individuals differed in the level of precision with which they can locate the target based on observing the agent's eyes.

The shape of the $P(x_c | \alpha_l, \alpha_r, \theta)$ distribution leads to a testable group-level prediction. As the pupil location varies, a fixed amount of uncertainty around the pupil angle corresponds to a varying degree of uncertainty in the estimated target location (see Figure \@ref(fig:fig2)B & C). When the agent directs their gaze toward the very left or right side, the distribution around the pupil angle from which participants sample is comparatively wider than when the agent gazes centrally to the ground in front of them. For illustrative purposes, imagine a similar phenomenon: pointing a torch light to a flat surface on the ground. When one points the light cone directly at the surface, the light beam is concentrated in a clearly defined, small, symmetric area. When one points the light cone further away from oneself (shining at an angle), the light from one half of the cone must travel further to reach the surface than the light from the other half, resulting in an asymmetric light pattern. As the angle increases, the light is spread over a wider area, and the surface is illuminated less evenly. Consequently, for the same $\sigma_v$, the further out a target coordinate lies, the wider and less symmetric the distribution. This increases both the variance and the bias in a participant's estimate of the agent's attentional focus, resulting in a decreased performance in the task. As $\sigma_v$ decreases and the cone narrows, the extent to which performance varies at different angles decreases. Therefore, our gaze model predicted that our trials vary in difficulty: participants should be more imprecise in locating the target the further out it lands, resulting in a U-shaped pattern.[^4] If our data matched the pattern of this model prediction, this could act as evidence for the gaze model. Therefore, our gaze model provides a quantitative theory of gaze understanding with testable predictions. In the following, we tested these predictions in children and adults.

@@

Our goal was to describe the inferential process of gaze understanding. Our analysis focused on the child sample. We quantified how well our gaze model explained the gaze understanding process in two ways. First, we used correlations to quantify how well the gaze model is able to recover the data. For this, we aggregated the model predictions and the data for each target bin and age group (3-, 4-, 5-years-olds, adults), and computed the coorrelation between predictions and the data. Second, we compared the predictions of our gaze model to two - very simple - alternative models that assume that participants do not rely on the agent's gaze at all: a random guessing model and a center bias model. The random guessing model assumed that participants randomly clicked on the screen and was implemented as sampling from a uniform distribution over all possible coordinates, $\mathcal{U}(0, 1920)$. The center bias model assumed that participants always click at the screen center and was implemented as sampling from a Normal distribution with the the center of the screen as the mean, and one balloon width as the variance, $\mathcal{N}(960, 160)$. Note that the center bias model also predicted that imprecision should be higher for targets further out on the screen. However, compared to the gaze model, the center bias model predicted a very steep effect towards the sides, resulting in a V-shaped pattern (i.e., imprecision as the distance between the target location and the screen center). All cognitive models were implemented in WebPPL [@goodman2014design].

@@

```{r gazemodel_results}
gazemodel_comparison_kids <- readRDS("../data/gazemodel-kids-comparison-bf.rds")
gazemodel_comparison_adults <- readRDS("../data/gazemodel-adults-comparison-bf.rds")
gazemodel_corrs <- readRDS("../saves/gazemodel_correlations.rds")
gazemodel_prior <- readRDS("../saves/gazemodel_prior.rds")
```

The age effects in Study 2 largely replicated those of Study 1. When comparing performance of the three- to five-year-olds in Study 1 and 2, we see a rapid improvement in gaze understanding in early childhood (see Figure \@ref(fig:fig1) inset and \@ref(fig:fig3)A). The gaze model's prior over potential target locations assumed that participants would be skewed to click towards the screen center. For the children, we estimated the width of the prior's distribution based on the participants' age. Interestingly, we found that the differential age effects on the center bias prior were minor (intercept = `r gazemodel_prior$mode[gazemodel_prior$type == "intercept"]`, 95%HDI [`r gazemodel_prior$lci[gazemodel_prior$type == "intercept"]`; `r gazemodel_prior$uci[gazemodel_prior$type == "intercept"]`]; slope = `r gazemodel_prior$mode[gazemodel_prior$type == "slope"]`, 95%HDI [`r gazemodel_prior$lci[gazemodel_prior$type == "slope"]`; `r gazemodel_prior$uci[gazemodel_prior$type == "slope"]`]). The slope of the prior indicated that older children were only marginally less drawn towards the screen center (i.e., marginally wider distribution) compared to the younger children.
